Verifying the \shaThree\ Implementation from \openssl\ with the Software Analysis Workbench

### 2022-05-13 paper/introduction.tex
Fix minor typos and make word changes

This paper discusses an automated proof of the \emph{Secure Hash Algorithm 3} (\shaThree) in \openssl\ using the \emph{Software Analysis Workbench} by Galois.
\shaThree\ is the latest algorithm for computing \emph{digests}, or hashes, from data to be published by the \emph{National Institute of Standards and Technology} (\nist).
The algorithm is the successor to \shaTwo.
Its intent is to address weaknesses in \shaTwo, and it departs from its predecessors in that it is based on the \keccak\ family of cryptographic primitives \cite{fips202}.

@@

The \emph{Software Analysis Workbench} (\saw) is a formal verification tool to prove equivalence between \sawcore\ models. \sawcore\ models are defined in the \cryptol\ functional language or extracted from C or Java with \emph{symbolic execution}.
\cryptol\ itself is designed to specify cryptographic primitives at the bit level.
\saw\ has been shown capable in proving implementations of cryptographic primitives equivalent to hardware and software implementations \cite{crypt-hi,hard-soft,design-verif}.
Recent work emphasizes not just the utility of \saw, but the importance of modular decomposition of cryptographic primitives in their definition as that decomposition lends itself to efficient verification in \saw\ \cite{nfm-us}.

>>>>

This paper discusses an automated proof of the \emph{Secure Hash Algorithm 3} (\shaThree) in \openssl\ using the \emph{Software Analysis Workbench} (\saw) by Galois.
\shaThree\ is the latest algorithm for computing \emph{digests}, or hashes, from data to be published by the \emph{National Institute of Standards and Technology} (\nist).
The algorithm is the successor to \shaTwo.
Its intent is to address weaknesses in \shaTwo, and it departs from its predecessors in that it is based on the \keccak\ family of cryptographic primitives \cite{fips202}.

@@

\saw\ is a formal verification tool to prove equivalence between \sawcore\ models. \sawcore\ models are defined in the \cryptol\ functional language or extracted from C or Java with \emph{symbolic execution}.
\cryptol\ itself is designed to specify cryptographic primitives at the bit level.
\saw\ has been shown capable in proving implementations of cryptographic primitives equivalent to hardware and software implementations \cite{crypt-hi,hard-soft,design-verif}.
Recent work emphasizes not just the utility of \saw, but the importance of modular decomposition of cryptographic primitives in their definition as that decomposition lends itself to efficient verification in \saw\ \cite{nfm-us}.

### 2022-05-13 paper/proof.tex
fixed minor typos
Fix minor typos and make word changes

The goal of the work presented in this paper is to prove the \openssl\ implementation of \shaThree\ matches the \fips\ publication by \nist.
\figref{fig:proofStructure} illustrates the proof strategy.
The proof begins on the left of the figure with the box labeled \emph{NIST Standard}. 
This box represents the \fips\ standard that defines the \shaThree\ algorithm with English prose and math.
The standard also provides several test inputs with the corresponding expected digests.
The first step of the proof reproduces the \fips\ \shaThree\ description in \cryptol.
That is the box to the right of the \emph{NIST Standard} box labeled \emph{Spec}.

@@

The syntactic disconnect between the quantification and looping in the published standard and the functional nature of the \cryptol\ specification is bridged with a library that hides list comprehensions with appropriately named functions.
The final specification using the library has a more direct correspondence to the published standard that is more obviously visually certified.
It size is around $200$ \emph{lines of code} (LOC) in \cryptol.

@@

The extracted \sawcore\ from the \openssl\ C implementation of \shaThree\ is too big and complex for \saw\ to reason about directly.
Adding to the complexity of the equivalence proof is that the \openssl\ implementation changes the meaning of the state $S$ in the algorithm so that it no longer directly matches the meaning defined in the published standard.
It also changes the computation of the inner \keccak\ functions.
These changes effectively reorder the matrix definition of the state to be more amenable to memory, and optimize the different functions in \keccak\ to be more efficient when operating on the state.
This includes using pre-computed constants for $\iota$.
The changes are significant enough to make it extremely difficult to argue manually that the computation implemented in \openssl\ matches that defined in the standard.

@@

\tableref{table:stats} is a summary of the cost of the verification in running time and LOCs. These are from fairly banal hardware seen in any laptop.
The \cryptol\ specification of the \fips\ standard relies on visual inspection and test vectors for equivalence having to look at around 200 LOC.
The equivalence between the \cryptol\ specification of the standard and the \cryptol\ specification of the \openssl\ specification that reorders the state array is limited to the 256 bit digest on input sizes from $0$ bits to $1,087$ bits and takes around 60 hours of computation time.
That is a proof that considers all of \algoref{alg:sha3}.
The equivalence between the \openssl\ \cryptol\ specification that reorders the state array and the actual C implementation is limited to the \keccak\ algorithm only.
That is a proof that only considers \lineref{line:keccak} in \algoref{alg:sha3} but holds for any state and is independent of the digest size and rate.
The cost of proving the equivalence for each of the overrides is trivial with the exception of $\iota$.
$\iota$ is more costly because it uses a lookup table in the C code so every entry in the table had to be proved out (see \secref{sec:openssl}).
The final cost of proving the equivalence in \keccak\ with the overrides is less than 15 minutes with the \texttt{yices} solver.

>>>>

The goal of the work presented in this paper is to prove the \openssl\ implementation of \shaThree\ matches the \fips\ specification by \nist.
\figref{fig:proofStructure} illustrates the proof strategy.
The proof begins on the left of the figure with the box labeled \emph{NIST Standard}. 
This box represents the \fips\ specification that defines the \shaThree\ algorithm with English prose and math.
The specification also provides several test inputs with the corresponding expected digests.
The first step of the proof reproduces the \fips\ \shaThree\ description in \cryptol.
That is the box to the right of the \emph{NIST Standard} box labeled \emph{Spec}.

@@

The syntactic disconnect between the quantification and looping in the published standard and the functional nature of the \cryptol\ specification is bridged with a library that hides list comprehensions with appropriately named functions.
The final specification using the library has a more direct correspondence to the published standard that is more obviously visually certified.
Its size is around $200$ \emph{lines of code} (LOC) in \cryptol.

@@

The extracted \sawcore\ from the \openssl\ C implementation of \shaThree\ is too big and complex for \saw\ to reason about directly.
Adding to the complexity of the equivalence proof is that the \openssl\ implementation changes the meaning of the state $S$ in the algorithm so that it no longer directly matches the meaning defined in the published standard.
It also changes the computation of the inner \keccak\ functions.
These changes effectively reorder the matrix definition of the state to be more amenable to memory and optimize the different functions in \keccak\ to be more efficient when operating on the state.
This includes using pre-computed constants for $\iota$.
The changes are significant enough to make it extremely difficult to argue manually that the computation implemented in \openssl\ matches that defined in the standard.

@@

\tableref{table:stats} is a summary of the cost of the verification in running time and LOCs. These are from fairly banal hardware seen in any laptop.
The \cryptol\ specification of the \fips\ standard relies on visual inspection and test vectors for equivalence having to look at around 200 LOC.
The equivalence between the \cryptol\ specification of the standard and the \cryptol\ specification of the \openssl\ specification that reorders the state array is limited to the 256 bit digest on input sizes from $0$ bits to $1,088$ bits and takes around 60 hours of computation time.
That is a proof that considers all of \algoref{alg:sha3}.
The equivalence between the \openssl\ \cryptol\ specification that reorders the state array and the actual C implementation is limited to the \keccak\ algorithm only.
That is a proof that only considers \lineref{line:keccak} in \algoref{alg:sha3} but holds for any state and is independent of the digest size and rate.
The cost of proving the equivalence for each of the overrides is trivial with the exception of $\iota$.
$\iota$ is more costly because it uses a lookup table in the C code so every entry in the table had to be proved out (see \secref{sec:openssl}).
The final cost of proving the equivalence in \keccak\ with the overrides is less than 15 minutes with the \texttt{yices} solver.

### 2022-05-13 paper/sha-3.tex
Fix minor typos and make word changes

\begin{algorithm}[t]
  \caption{\shaThree\ sponge construction with \keccak}\label{alg:sha3}
  \begin{algorithmic}[1]
    \Require $n \geq 0$ and $(d,r) \in \{(224, 1152)\ (256, 1088)\ (384, 832)\ (512, 576)\}$
    \Ensure $D =$ \shaThree(N,d,r)
    \State $p_0, \ldots, p_m = \mathtt{pad}(N, r)$\label{line:pad}
    \State $A = 0_0, \ldots, 0_{1599}$\label{line:init}
    \For{$p_i = p_0, \dots, p_m$}
      \State $A = A\ \mathbf{xor}\ \mathtt{zeroExtend}(p_i, a-r)$\label{line:absorb}
      \State $A =$ \keccak$(A)$\label{line:keccak}
    \EndFor
    \State $D = \mathtt{squeeze(A,d)}$\label{line:squeeze}
  \end{algorithmic}
\end{algorithm}

\shaThree\ uses \emph{sponge construction} to generate hashes as shown in \algoref{alg:sha3} (see \nist\ \fips\ \cite{fips202}).
Sponge construction \emph{absorbs} data into an internal state $A$ at a rate of $r$ bits per cycle (\lineref{line:absorb} and \lineref{line:keccak}), transforming the state on each cycle, and once all the data is absorbed, then it \emph{squeezes} out a final $d$-bit digest (\lineref{line:squeeze}).
The transformation function, \keccak-f[1600], simply \keccak\ in the rest of this presentation, defines the size of the state $A$ to be $a = 1600$ bits, which is initialized to 0 (\lineref{line:init}).

>>>>

\begin{algorithm}[t]
  \caption{\shaThree\ sponge construction with \keccak}\label{alg:sha3}
  \begin{algorithmic}[1]
    \Require $n \geq 0$ and $(d,r) \in \{(224, 1152)\ (256, 1088)\ (384, 832)\ (512, 576)\}$
    \Ensure $D =$ \shaThree(N,d,r)
    \State $p_0, \ldots, p_m = \mathtt{pad}(N, r)$\label{line:pad}
    \State $A = 0_0, \ldots, 0_{1599}$\label{line:initkeccak}
    \For{$p_i = p_0, \dots, p_m$}
      \State $A = A\ \mathbf{xor}\ \mathtt{zeroExtend}(p_i, a-r)$\label{line:absorb}
      \State $A =$ \keccak$(A)$\label{line:keccak}
    \EndFor
    \State $D = \mathtt{squeeze(A,d)}$\label{line:squeeze}
  \end{algorithmic}
\end{algorithm}

\shaThree\ uses \emph{sponge construction} to generate hashes as shown in \algoref{alg:sha3} (see \nist\ \fips\ \cite{fips202}).
Sponge construction \emph{absorbs} data into an internal state $A$ at a rate of $r$ bits per cycle (\lineref{line:absorb} and \lineref{line:keccak}), transforming the state on each cycle, and once all the data is absorbed, then it \emph{squeezes} out a final $d$-bit digest (\lineref{line:squeeze}).
The transformation function, \keccak-f[1600], simply \keccak\ in the rest of this presentation, defines the size of the state $A$ to be $a = 1600$ bits, which is initialized to 0 (\lineref{line:initkeccak}).
